---
title: "Programming in the R Language"
subtitle: "MY 470, Week 9: Advanced R Programming"
author: "Kenneth Benoit"
date: "2017-11-20"
output: 
  ioslides_presentation:
    widescreen: true
---

# (More) Advanced R Programming
    
## `matrix`

A matrix is just a specially attributed vector.

- a vector with dimension attributes, where all elements are of the same type

```{r}
my_matrix <- matrix(data = 1:100, nrow = 10, ncol = 10)
my_matrix
```

## `data.frame`

A `data.frame` is a matrix-like R object in which the columns can be of different types.  It is actually a special form of a **list**.

```{r}
char_vec <- c("apple", "pear", "plumb", "pineapple", "strawberry")
logical_vec <- c(TRUE, FALSE, FALSE, TRUE, FALSE)

my_data_frame <- data.frame(numbers = c(5, 4, 2, 100, 7.65),
                            fruits = char_vec, 
                            logical = logical_vec)
my_data_frame
```

## Beware: `stringsAsFactors = TRUE` by default

```{r}
str(my_data_frame)
```

## How to correct this:

```{r}
my_data_frame <- data.frame(numbers = c(5, 4, 2, 100, 7.65),
                            fruits = char_vec, 
                            logical = logical_vec,
                            stringsAsFactors = FALSE)
str(my_data_frame)
```


## `matrix` and `data.frame` subsetting

To subset a `matrix` or `data.frame`, you need to specify both rows and columns:

```{r}
my_matrix[1:3, 1:3]

my_data_frame[1, ]
```

## `matrix` and `data.frame` subsetting

We can also subset to remove rows or columns that we do not want to see by using the `-` operator applied to the `c` function:

```{r}
my_matrix[-c(1:3), -c(1:3)]
```

In the code, `1:3` creates a vector of the integers 1, 2 and 3, and the `-` operator negates these. We wrap the vector in the `c` function so that `-` applies to each element, and not just the first element.

## Index addressing for vector and list elements can be done in different ways
```{r}
x <- c(city = "London", university = "LSE", module = "my470")
```

### by integer index value, e.g.
```{r}
x[2]
x[c(1, 3)]
```

----

```{r}
x <- c(city = "London", university = "LSE", module = "my470")
```

### by the `TRUE` elements in a logical vector, e.g. `x[c(TRUE, FALSE, TRUE)]`
```{r}
x[c(TRUE, FALSE, TRUE)]
```

----

```{r}
x <- c(city = "London", university = "LSE", module = "my470")
```

### by character name, e.g. `x["city"]`
```{r}
x["city"]
```

## generalizes to matrix-like objects, through row and column names

```{r}
x <- matrix(1:6, nrow = 2, 
            dimnames = list(c("one", "two"), c("a", "b", "c")))
x
x[, 2]     # all rows, second column
```

----
```{r}
x[1, 2:3]  # first row, cols 2 through 3
x[c("one"), c("b", "c")]  # same
```


## `list`

A `list` is a collection of any set of object types

```{r}
my_list <- list(something = my_data_frame$num_vec, 
                another_thing = my_matrix[1:3,1:3], 
                something_else = "ken")
my_list
```


## How to index list elements in R

### Using `[`
```{r}
my_list["something_else"]
my_list[3]
```

----

### Using `[[`
```{r}
my_list[["something"]]
my_list[[1]]
```

----

### Using `$`
```{r}
my_list$another_thing
```

(Does not allow multiple elements to be indexed in one command)

## `functions`

R makes extensive use of functions, which all have the same basic structure.

```{r, eval = FALSE}
function_name(argument_one, argument_two, ...)
```

Where 

- `function_name` is the name of the function
- `argument_one` is the first argument passed to the function
- `argument_two` is the second argument passed to the function

## using function arguments

>- when a function is not assigned a _default value_, then it is mandatory
>- when a function has a default, this is used but can be overridden
>- it is not necessary to specify the names of the arguments, although it is best to do so except for the first or possibly second arguments

## `function` example

Let's consider the `mean()` function. This function takes two main arguments:

```{r, eval = FALSE}
mean(x, na.rm = FALSE)
```

Where `x` is a numeric vector, and `na.rm` is a logical value that indicates whether we'd like to remove missing values (`NA`).

```{r, eval = TRUE}
vec <- c(1, 2, 3, 4, 5)
mean(x = vec, na.rm = FALSE)
```

```{r, eval = TRUE}
vec <- c(1, 2, 3, NA, 5)
mean(x = vec, na.rm = TRUE)
```

## `function` example

We can also perform calculations on the output of a function:
```{r, eval = TRUE}
vec <- 1:5
mean(vec) * 3
```

Which means that we can also have nested functions:
```{r, eval = TRUE}
sqrt(mean(vec))
```

We can also assign the output of any function to a new object for use later:
```{r, eval = TRUE}
sqrt_mean_of_vec <- sqrt(mean(vec))
```

## User defined `functions`

Functions are also objects, and we can create our own. We define a function as follows:
    
```{r}
my_addition_function <- function(a = 10, b) {
    a + b
}

my_addition_function(a = 5, b = 50)
my_addition_function(3, 4)
my_addition_function(b = 100)
```

## Variables in functions have local scope

```{r}
my_demo_function <- function(a = 10) {
    a <- a * 2
    a
}

a <- 1
my_demo_function(a = 20)
a
```


# Reading data into R

## Reading data into R (.csv)

```{r, eval = FALSE}
my_data <- read.csv(file = "my_file.csv")
```
    
- `my_data` is an R data.frame object (you could call this anything)
- `my_file.csv` is a .csv file with your data
- `<-` is the assignment operator
- In order for R to access `my_file.csv`, it will have to be saved in your current working directory
  - Use `get_wd()` to check your current working directory
  - Use `set_wd()` to change your current working directory
- Might need to use the `stringsAsFactors = FALSE` argument


## (creating some fake data)
```{r, eval = TRUE}
n <- 1000
x <- rnorm(n)
z <- runif(n)
g <- sample(letters[1:6], n, replace = T)
beta <- 0.5
beta2 <- 0.3
beta3 <- -0.4
alpha <- 0.3
eps <- rnorm(n,sd = 1)
y <- alpha + beta*x + beta2*z + beta3 *(x*z) + eps
y_bin <- as.numeric(y > median(y))
my_data <- data.frame(x = x, y = y, z = z, g = g)
```
    

# Plots and graphs

## Introduction

- Plots are one of the great strength f R.
- There are two main frameworks for plotting
- Base R graphics
- **ggplot**

## Base R plots
    
The basic plotting syntax is very simple. `plot(x_var, y_var)` will give you a scatter plot:
```{r, fig.align = "center", fig.width = 6}
plot(my_data$x, my_data$y)
```

This can be substantially improved!

## Base R plots

The plot function takes a number of arguments (`?plot` for a full list). The fewer you specify, the uglier your plot:

```{r, eval=FALSE}
plot(x = my_data$x, y = my_data$y, 
     xlab = "X variable",            # x axis label
     ylab = "Y variable",            # y axis label 
     main = "My spilled bubble gum", # main title
     pch = 19,                       # solid points
     cex = 0.5,                      # smaller points
     bty = "n",                      # remove surrounding box
     col = as.factor(my_data$g)      # colour by grouping variable
)
```

## Base R plots

```{r, echo=FALSE, fig.align="center"}
plot(my_data$x, my_data$y, 
     xlab = "X variable", 
     ylab = "Y variable", 
     main = "My spilled bubble gum",
     pch = 19, # Solid points
     cex = 0.5, # Smaller points
     bty = "n", # Remove surrounding box
     col = as.factor(my_data$g) # Convert grouping variable to factor, colour by this
)
```

## Base R plots

The default behaviour of `plot()` depends on the type of input variables for the `x` and `y` arguments. If `x` is a factor variable, and `y` is numeric, then R will produce a boxplot:

```{r, eval=FALSE, fig.align="center"}
plot(x = my_data$g, y = my_data$x)
```

## Base R plots

```{r, echo=FALSE, fig.align="center"}
plot(x = my_data$g, y = my_data$x)
```

## ggplot

Also popular is the **ggplot2** library. This is a separate package (i.e. it is not a part of the base R environment) but is very widely used.

- Based on the "Grammar of Graphics"" data visualisation scheme
- Graphs are broken into scales and layers
- Has slightly idiosyncratic language style!

## ggplot

Let's recreate the previous scatter plot using `ggplot`:
    
```{r, eval=FALSE, fig.align="center"}
library("ggplot2"")
ggplot(data = my_data, aes(x= x, y= y, col = g)) + 
    geom_point() +
    xlab("X variable")+
    ylab("Y variable")+
    ggtitle("ggplot title")+
    theme(panel.background = element_rect("white"))
```

## ggplot

```{r, echo=FALSE, fig.align="center"}
library("ggplot2")
ggplot(data = my_data, aes(x= x, y= y, col = g)) + 
    geom_point() +
    xlab("X variable")+
    ylab("Y variable")+
    ggtitle("ggplot title")+
    theme(panel.background = element_rect("white"))
```

## ggplot

One nice feature of `ggplot` is that it is very easy to create facet plots:
    
```{r, eval=FALSE, fig.align="center"}
library("ggplot2")
ggplot(data = my_data, aes(x= x, y= y, col = g)) + 
    geom_point() +
    xlab("X variable")+
    ylab("Y variable")+
    ggtitle("ggplot facet title")+
    theme(panel.background = element_rect("white"))+
    facet_wrap(~ g)
```

## ggplot

```{r, echo=FALSE, fig.align="center"}
library("ggplot2")
ggplot(data = my_data, aes(x= x, y= y, col = g)) + 
    geom_point() +
    xlab("X variable")+
    ylab("Y variable")+
    ggtitle("ggplot facet title")+
    theme(panel.background = element_rect("white"))+
    facet_wrap(~ g)
```

## Other examples of pretty R graphics
```{r fig.width=10, fig.height=5,echo=FALSE}
library(png)
library(grid)
img <- readPNG("figs/spark_plot_new_trans.png")
grid.raster(img)
```


## Other examples of pretty R graphics

```{r fig.width=10, fig.height=5,echo=FALSE}
library(png)
library(grid)
img <- readPNG("figs/PartyLeaderSatisfaction.png")
grid.raster(img)
```

## Linear Regression Models

Linear regression models in R are implemented using the `lm()` function.

```{r}
my_lm <- lm(formula = y ~ x, data = my_data)
```

The `formula` argument is the specification of the model, and the `data` argument is the data on which you would like the model to be estimated.

```{r}
print(my_lm)
```

## `lm`

We can specify multivariate models:
    
```{r}
my_lm_multi <- lm(formula = y ~ x + z, data = my_data)
```

Interaction models:
    
```{r}
my_lm_interact <- lm(formula = y ~ x * z, data = my_data)
```

Fixed-effect models:
    
```{r}
my_lm_fe <- lm(formula = y ~ x + g, data = my_data)
```

And many more!
    
## `lm`
    
The output of the `lm` function is a long list of interesting output. 

When we call `print(saved_model)`, we are presented with the estimated coefficients, and nothing else.

For some more information of the estimated model, use `summary(saved_model)`:
    
```{r, eval = FALSE}
my_lm_summary <- summary(my_lm)
print(my_lm_summary)
```

## `lm`


```{r, echo = F, size="small"}
my_lm_summary <- summary(my_lm)
cat(paste0(capture.output(my_lm_summary)[-c(1:4)], collapse = "\n"))

```

## `lm`

As with any other function, `summary(saved_model)` returns an object. Here, it is a list. What is saved as the output of this function?
    
```{r}
names(my_lm_summary)
```

If we want to extract other information of interest from the fitted model object, we can use the `$` operator to do so:
    
```{r}
print(my_lm_summary$r.squared)
```

## `lm`

Accessing elements from saved models can be very helpful in making comparisons across models:
    
```{r}
my_lm_r2 <- summary(my_lm)$r.squared
my_lm_multi_r2 <- summary(my_lm_multi)$r.squared
my_lm_interact_r2 <- summary(my_lm_interact)$r.squared

r2.compare <- data.frame(
    model = c("bivariate", "multivariate", "interaction"), 
    r.squared = c(my_lm_r2, 
                  my_lm_multi_r2, 
                  my_lm_interact_r2))
```

## `lm`

We can print the values:
    
```{r}
print(r2.compare)
```

Or we can plot them:
```{r, eval = FALSE}
ggplot(r2.compare, aes(x = model, y = r.squared))+
    geom_point(size = 4)+
    ggtitle("R^2 Comparison")
```

## `lm`

```{r, echo= F, fig.align="center"}
ggplot(r2.compare, aes(x = model, y = r.squared))+
    geom_point(size = 4)+
    ggtitle("R^2 Comparison")+
    theme(panel.background = element_rect("white"))
```

## `lm` diagnostics

There are a number of functions that are helpful in producing model diagnostics:
    
- `residuals(saved_model)` extracts the residuals from a fitted model
- `coefficients(saved_model)` extracts coefficients
- `fitted(saved_model)` extracts fitted values
- `plot(saved_model)` is a convenience function for producing a number of useful diagnostics plots

## `lm` residual plot

We can easily plot the residuals from a fitted model against an explanatory variable of interest:
    
```{r, eval = FALSE}
par(mfrow = c(1,2)) # Divide the plotting region into 1 row and 2 cols
plot(x = my_data$x, y = residuals(my_lm_multi), 
     xlab = "x", ylab = "residuals",  # axis labels
     pch = 19,                        # filled circles
     col="grey",                      # mhange colour
     cex = 0.5)                       # make point size smaller
abline(h = 0)                         # add a horizontal line with y-intercept of 0
plot(x = my_data$z, y = residuals(my_lm_multi), 
     xlab = "z", ylab = "residuals",
     pch = 19, col = "grey", cex = 0.5) 
abline(h = 0)                         
```

## `lm` residual plot

```{r, echo=F, fig.align="center"}
par(mfrow = c(1,2)) # Divide the plotting region into 1 row and 2 cols
plot(my_data$x, residuals(my_lm_multi), 
     xlab = "x", ylab = "residuals",
     pch = 19, col="grey", cex = 0.5)
plot(my_data$z, residuals(my_lm_multi), 
     xlab = "z", ylab = "residuals",
     pch = 19, col = "grey", cex = 0.5)
```

# Non-Linear Regression Models

## `glm`

To estimate a range of non-linear models, the `glm` function is particularly helpful.

First, let us transform our outcome variable from a continuous measure to a binary measure:
    
```{r, eval = TRUE}
my_data$y_bin <- as.numeric(my_data$y > median(my_data$y))
table(my_data$y_bin)
```

## `glm`

Now we can estimate our model:
    
```{r}
my_logit <- glm(formula = y_bin ~ x + z, data = my_data, family = "binomial")
```

Where:
    
- `formula` is the model specification
- `data` is the data
- `family` is a description of the error distribution and link function to be used
- `binomial`, `poisson`, `gaussian` etc...

## `glm`

```{r, eval = FALSE}
summary(my_logit)
```

```{r, echo = FALSE}
cat(paste0(capture.output(summary(my_logit))[-c(1:4)], collapse = "\n"))
```

## `glm`

OK, but no-one actually thinks in terms of log-odds, so let's translate that into something more meaningful.

```{r, message=FALSE}
my_logit_OR <- exp(cbind(OR = coef(my_logit), confint(my_logit)))
```

- `coef` extracts the coefficients
- `confint` extracts the confidence intervals
- `cbind` binds the vectors together as separate *c*olumns
- `exp` exponentiates the log-odds ratios

```{r, message=F, echo=TRUE}
round(my_logit_OR, digits = 4)
```

## `glm`

Almost all of the convenience functions that we used for `lm` are also applicable to `glm` models:

```{r, eval=FALSE}
summary(my_logit)
plot(my_logit)
residuals(my_logit)
coefficients(my_logit)
fitted(my_logit)
```

## Other models

There are a number of external packages that can make fitting other model types (relatively) straightforward:

- `lmer4` - Linear, generalised linear, and nonlinear mixed models
- `mcgv` - generalised additive models
- `survival` - survival analysis
- `glmnet` - lasso and elastic net regression models
- `randomForest` - random forest models from machine learning
- `rjags` and `rstan` - Bayesian models

To use a package that is not a part of the base R installation, use:

```{r, eval = FALSE}
install.packages("survival")
library(survival)
```

## `predict`

We can retreive the fitted values from the model using `fitted()`, but we may be interested in calculating predicted values for arbitrary levels of our covariates.

```{r}
sim_data <- data.frame(x = c(0, 1))
y_hat <- predict(object = my_lm, newdata = sim_data)
y_hat
```

Here, I am creating a `data.frame` with two observations of one variable (`x`).

I am then using the `predict` function, where

- `object = my_lm` tells R the model object for which prediction is desired
- `newdata = sim_data` tells R the values for which I would like predictions

## `predict`

We can use the same syntax to retreive predictions for (marginally) more interesting models:

```{r}

sim_data <- data.frame(x = c(0, 0, 1, 1), z = c(0, 1, 0, 1))
sim_data
y_hat <- predict(my_lm_multi, newdata = sim_data)
y_hat
```

## `predict`

This can be especially useful when trying to visualise interactive models:
```{r, eval = TRUE}

sim_data_z0 <- data.frame(x = seq(from = -2, to = 2, by = 0.01), z = 0)
sim_data_z1 <- data.frame(x = seq(from = -2, to = 2, by = 0.01), z = 1)
y_hat_z0 <- predict(my_lm_interact, newdata = sim_data_z0)
y_hat_z1 <- predict(my_lm_interact, newdata = sim_data_z1)
```

- `seq` generates a regular sequences `from` one value `to` another value `by` given increments
- I am creating two `data.frame`s for prediction, in both cases varying the value of `x`, but first setting `z` to 0, and then setting `z` to 1

## `predict`

```{r, eval = FALSE}
# Create a plot of the data
plot(my_data$x, my_data$y, cex = 0.5, pch = 19, 
col = "gray", bty = "n",
xlab = "X", ylab = "Y", 
main = "Fitted values for sim_data")

# Add a prediction line for z = 0
lines(x = sim_data_z0$x, y = y_hat_z0, lwd = 2)

# Add a prediction line for z = 1
lines(x = sim_data_z1$x, y = y_hat_z1, lwd = 2, col = "red")

# Add a legend
legend("bottomright", legend = c("z = 0", "z = 1"), col = c("black", "red"), lty = 1) 
```

## `predict`

```{r, echo = F, fig.align="center"}
par(mfrow = c(1,1))
plot(my_data$x, my_data$y, cex = 0.5, pch = 19, 
col = "gray", bty = "n",
xlab = "X", ylab = "Y", 
main = "Fitted values for sim_data")
lines(x = sim_data_z0$x, y = y_hat_z0, lwd = 2)
lines(x = sim_data_z1$x, y = y_hat_z1, lwd = 2, col = "red")
legend("bottomright", legend = c("z = 0", "z = 1"), col = c("black", "red"), lty = 1)
```

## Can we just use `margins`?

Yes! Thomas Leeper has created a nice port of STATA's `margins` command:
    
```{r, eval=FALSE}
library(margins)

cplot(my_lm_interact, x = "x", dx = "z", what = "effect")
```

`cplot` is equivalent to `marginsplot` in STATA. 

The specification above tell `R` that we want to plot the marginal effect of `z` on `y` across the range of `x`.

## Can we just use `margins`?

```{r, eval=TRUE, echo=FALSE, fig.align="center"}
library(margins)

cplot(my_lm_interact, x = "x", dx = "z", what = "effect")

```

## Other helpful functions

```{r, echo = T, eval = FALSE}
objects() / ls() # Which objects are currently loaded in my working environment?
rm()             # Remove objects from my current environment
save()           # Save R object(s) to disk
is.na()          # Is this object a missing value? Or, which elements in this vector are missing?
rnorm()          # Generate random numbers from a normal distribution
runif()          # Generate random numbers from a uniform distribution
```

## Other helpful packages

```{r, echo = T, eval = FALSE}
library(data.table)
library(dplyr)
library(plyr)
library(zoo)
library(reshape2)
library(shiny)
```