# Exercise One

# Data manipulation and exploration

## Reading data into R (.csv)

```{r, eval = F}

my_data <- read.csv(file = "my_file.csv")

```

- `my_data` is an R data.frame object (you could call this anything)
- `my_file.csv` is a .csv file with your data
- `<-` is the assignment operator
- In order for R to access `my_file.csv`, it will have to be saved in your current working directory
- Use `get_wd()` to check your current working directory
- Use `set_wd()` to change your current working directory

## Reading data into R (.dta)

```{r, eval = F}
install.packages("foreign") ## install the "foreign" package
library(foreign) ## Loads an additional package that deals with STATA files
my_data <- read.dta(file = "my_file.dta") 

```

```{r, eval = T, echo = F}
n <- 1000

x <- rnorm(n)
z <- runif(n)
g <- sample(letters[1:6], n, replace = T)
beta <- 0.5
beta2 <- 0.3
beta3 <- -0.4
alpha <- 0.3
eps <- rnorm(n,sd = 1)
y <- alpha + beta*x + beta2*z + beta3 *(x*z) + eps
y_bin <- as.numeric(y > median(y))

my_data <- data.frame(x = x, y = y, z = z, g = g)

```

- The `foreign` package is not automatically installed in R so we need to do that ourselves
- `my_data` is still an R data.frame object (you could still call this anything)
- `my_file.dta` is a .dta file with your data
- `<-` is still an assignment operator

## Descriptive statistics functions

R includes a number of helpful functions for exploring your data

```{r, eval = F}
View(my_data) # View data in spreadsheet-style
names(my_data) # Names of the variables in your data
head(my_data) # First six rows of the data
tail(my_data) # Last six rows of the data
str(my_data) # "Structure" of any R object
summary(my_data) # Summary statistics for each of the columns in the data
dim(my_data) # Dimensions of a matrix
range(my_data$my_variable) # Range of a numeric vector
quantile(my_data$my_variable) # Quantiles of a numeric vector
```

## `head`
```{r}
head(my_data) 
```

## `str`
```{r}
str(my_data) 
```

## `summary`
```{r}
summary(my_data) 
```

## `pairs`

```{r}
pairs(my_data)
```


## `range`

Suppose you would like to find the range of a variable in your `data.frame`. The following will not work:
    
```{r, error=T}
range(my_data) 
```

Right, we cannot take the range of the entire data! Some of the variables are not numeric, and therefore do not have a range...

## `range`

Instead, we need to access one variable. There are two ways to do this. 

- We can subset using square parenthesis, or
- We can use the `$` operator and access the variable of interest by name

```{r}
range(my_data[,1])
range(my_data$x) 

```
